Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 parallel.tex
Updated Section 12 (parallel performance)

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained on a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could be
either a core or a node, depending on details of the system.  Note
that the current version of \enzo\ is not threaded, although a hybrid
MPI + OpenMP version is currently under development.}.  
Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors, $N_p$.

Load balancing on levels other than the root level (i.e., grids for
which the level $l > 0$) is different, as the refined patches are not
generally uniformly distributed.  Grids on refined levels are first
placed on the same processor as their parent to minimize
communication; however, this generally does not result in a
well-balanced computational load.  Therefore, the code has a number of options for
load balancing the grids on a given level. Each grid is assigned an
estimated computational load, generally equal to the number of cells in the grid (which,
empirically, is a good estimate of computational cost).  The first
load-balancing option is to move a grid from the processor with the
highest computational load to the processor with the lowest load, with the proviso
that only grids with load factors less than half the difference
between the highest and lowest loaded processors will be moved.  This
continues until the load ratio between the most-to-least loaded
processor is below 1.05 or until no suitable grid can be found to
transfer.  A second load balancing option uses a space-filling Hilbert
curve to order the grids by their approximate spatial position.  Then,
once the grids have a specific one-dimensional ordering, we can divide
up the grids into $N_p$ groups (with the division taking place as
equally as possible).  Load balancing is done separately for each
level.  Clearly load balancing is most successful if there are
significantly more grids than processors; however, small grids are
less efficient (because of their ghost zones), and so the code uses a
simple heuristic in order to split up grids until there are of order
10 grids per processor.  This generally results in good load-balancing
while not producing grids that are wastefully small.

Communication between processors is done using a non-blocking
communication strategy that allows overlap of communication and
computation.  This can be done efficiently because each processor
retains a copy of the entire hierarchy of grids, except that grids
that do not `live' on a given processor only contain meta data
(essentially location and size of the grid; such grids are denoted as
`ghost' grids).  Replicating the hierarchy means that all
communication of data from one grid to another can be identified by
each processor independently.  The metadata for `ghost' grids are
quite small and so the extra memory required is generally not onerous
unless very large numbers of grids are used (more than a few hundred
thousand grids).  A schematic of this distribution is shown in
Figure~\ref{fig.amr_hierarchy}.

Data is transferred through a three-step procedure that takes
advantage of the capabilities of the MPI library: (i) as the code
progresses, and data is needed from another grid on another processor,
the receiving processor posts an MPI non-blocking receive indicating
that it is expecting data; this outstanding receive is recorded in a
table, (ii) the sending processor calls the MPI non-blocking send
function, and then finally (iii) the receiving processor, after it has
carried out all the computation it can, waits for any MPI message to
arrive.  Each message is coded so that it can be matched with the
appropriate receive posted in the first step, and based on that, the
appropriate routine is called to processes the data.  Step (iii) is
repeated until there are no outstanding receives.

@@

At each complete root grid timestep (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells updates per second per MPI process.  This
information is then 
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with further discussion of the
performance measurement implementation.

@@

It is advantageous to use \enzo\ in its ``unigrid'' (i.e.,
non-adaptive mesh) mode for a variety of problems, including
non-gravitating turbulence
\citep[e.g.,][]{2002ApJ...569L.127K,Kritsuk04, 2007ApJ...665..416K,
2009ASPC..406...15K}, the Lyman-$\alpha$ forest
\citep{2005MNRAS.361...70J,2009MNRAS.399.1934P}, or feedback of
metal-enriched gas from galaxies into the intergalactic medium
\citep{2004ApJ...601L.115N,2011ApJ...731....6S}.  Achieving good
scaling of the code in unigrid mode is relatively straightforward --
upon initialization, unigrid simulations are decomposed such that each
MPI process has a roughly equal subvolume (and thus an equal number of grid
cells), meaning that work is evenly distributed.  Communication
patterns for both the gravity solve (which uses a fast Fourier
transform) and the fluid solves (which transfer boundary information
between subvolumes) are predictable and straightforward, and
rebuilding of the grid hierarchy does not take place, removing a
substantial global operation and a great deal of communication.

@@

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this value has
been empirically determined to be roughly $128^3$ cells per core.  If
fewer cells per core are used, the CPU is essentially data-starved,
and poor scaling is observed due to computing units being idle while
waiting for information to be communicated from other processes (for,
e.g., boundary information or gravity solves).  Substantially larger
cell counts per core would in principle help scaling by reducing the
amount of inter-process communication needed, but larger cell counts
are typically impractical on most machines due to memory limits.

As a final point, we observe that scaling at larger core counts has
been measured, but only with an experimental hybrid-parallel (MPI +
OpenMP) version of \enzo.  Using this version, scaling comparable to
that shown in Figure~\ref{fig.uniscale} was seen on up to 98,304 cores
on the NICS Kraken XT5 (an upgraded version of the XT4 machine used
for the scaling study shown in the figure), using 2-8 OpenMP threads
per MPI process.  Hybrid parallelism has the potential to
substantially improve scaling by reducing the amount of communication
per grid tile (as described in the previous paragraph); however, the
optimal ratio of OpenMP threads per MPI task seems to vary
substantially between computational platforms and astrophysical
problems/required \enzo\ physics modules, and thus we hesitate to
provide strict guidelines here.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
\enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in \enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with fewer
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved more quickly, and the
opposite occurs with larger simulations.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per timestep is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late times.
Therefore, simply performing a few steps at the beginning of the
calculation does produce a good estimate of the required CPU time.

@@

The physical reason for the $f_{l+1}/f_l$ ratios found in the
cosmological simulations appears to be due to a combination of the
density structure of individual clouds, and the distribution of the
clouds themselves.  In particular, we note that for a $\rho \propto
r^{-2}$ density profile, the Lagrangian refinement criteria typically used in
such simulations produces an $f_{l+1}/f_l$ ratio of 1/8 for a single,
resolved cloud.  Of course, at some point we would resolve the flat
density of individual clouds and the ratio would climb, making further
refinement more costly; however, we are not yet in this regime in this
example.

@@

Many of today's leading supercomputers use a heterogeneous structure:
on a single node of a distributed-memory platform, a multi-core CPU is
often paired with one or more many-core accelerators.  One programming
model that has been shown to successfully take advantage of these
hardware accelerators is to run the serial component of the algorithm
on the CPU, and the vector-parallel part of the algorithm on the
many-core hardware accelerator. Due to the much higher computational
performance of the many-core accelerators, if a code can be ported to
effectively utilize this heterogeneous architecture, massive speedup
in simulation performance may be achieved.  This heterogeneous trend
is likely to continue in future supercomputers, due to the relatively
low energy needs of hardware accelerators as well as other
factors. Thus, to ensure \enzo's efficiency on future high-end
computational platforms, some of the most time-consuming parts of
\enzo\ have been ported to many-core architecture.  Since GPUs are
currently the most popular many-core accelerator, NVIDIA's CUDA C/C++
programming technology was chosen to port some \enzo\ modules to
NVIDIA GPUs.

@@

The key step in porting to many-core architectures such as the GPU is
exposing the massive parallelism inherent in the algorithm. Due to the
explicit, directionally-split stencil pattern of both the PPM and
Dedner MHD solvers, they are inherently massively parallel and thus
should be good fit for hardware acceleration. Both solvers essentially
contain basically two parts -- computation of fluxes, and cell update.
When computing fluxes, the basic procedure is to compute the flux at
each cell-interface given the inputs from neighboring cells. In the
cell update part, the fundamental computation is updating the
cell-centered values using the previously-computed fluxes. In the CPU
serial code, a loop over the grid is used, where the loop body
contains these basic computations. In both parts,
algorithmically-different loop iterations are completely independent
of each other. Thus, the parallelization scheme is naturally mapping
one GPU thread to one iteration of the loop. However, the original CPU
code, which operates on each grid as a serial process, does not
completely expose this parallelism as some small temporary arrays are
reused among loop iterations.  This re-use of arrays introduces data
dependency among loop iterations, which is undesirable for GPU
parallelization.  Because of this re-use of arrays, the main change in
porting the serial CPU solvers to allow massively parallel GPU
computation was replacing those shared temporary arrays by larger
temporary arrays that are not shared among loop iterations. This
change exposed the massive parallelism in the algorithm, which can
then be accelerated in a straightforward manner using CUDA.

To illustrate the speedup provided by porting two solvers to GPUs, we
show the results of a weak scaling test of driven MHD turbulence on a
uniform mesh in Figure~\ref{fig:gpu_scaling}.  This particular problem
type contains no physics other than solving the equations of ideal
MHD, and thus is representative of the type of calculation than can
benefit from the GPU-optimized solvers in \enzo.  In this scaling
test, we use the Dedner MHD solver, which runs on both CPUs and GPUs
and has been tuned to maximize performance on both platforms.  The
benchmarking platform is a Cray XK6 system with a Gemini
interconnect. Each node has a 16-core AMD Opteron 6272 CPU and a
single NVIDIA Tesla K20 GPU, which has 2,496 CUDA cores and a maximum
theoretical speed of 1.17 Tflops when used in double precision.  For
the CPU run, all the 16 cores in each node are used by launching 16
MPI processes per node. For the GPU run, 16 MPI processes are also
launched per node, and all processes on this node share the single GPU
on that node. This can work because NVIDIA's MPS (Multi-Process
Service) technology allows multiple processes to concurrently use the
same GPU.  A series of weak scaling tests were run, varying from 1 to
8 nodes (16 to 128 MPI processes), with each node containing a cube of
$256^3$ cells (so, a 4-node computation would have a domain that is
$512 \times 512 \times 256$ cells).  Figure~\ref{fig:gpu_scaling}
displays the scaling results in terms of cell updates per second per
node -- as a result, ideal weak scaling is a horizontal line.  In
general, the simulations using the GPU-accelerated MHD solver perform
roughly 5 times better overall on this system than the equivalent
CPU-only calculation run on the same number of nodes.  We note,
however, that this somewhat mis-represents the speedup obtained by
porting the MHD code to GPUs -- in the GPU simulation case, the 16 CPU
cores on each node are effectively idle, as they are used only for
timestep calculation and boundary condition transmission (which are
both computationally inexpensive compared to the fluid solve).  This
means that, effectively, the GPU simulation is 80 times faster than a
single CPU core, so a user would likely see much greater effective
speedup on a system where the ratio of CPU cores to GPUs is lower.

>>>>

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained on a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could be
either a core or a node, depending on details of the system.  Note
that the current version of \enzo\ is not threaded, although a hybrid
MPI + OpenMP version is under development.}.  
Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors, $N_p$.

Load balancing on levels other than the root level (i.e., grids for
which the level $l > 0$) is different, as the refined patches are not
generally uniformly distributed.  Grids on refined levels are first
placed on the same processor as their parent to minimize
communication; however, this generally does not result in a
well-balanced computational load.  Therefore, the code has a number of options for
load balancing the grids on a given level. Each grid is assigned an
estimated computational load, generally equal to the number of cells in the grid (which,
empirically, is a good estimate of computational cost).  The first
load-balancing option is to move a grid from the processor with the
highest computational load to the processor with the lowest load, with the proviso
that only grids with load factors less than half the difference
between the highest and lowest loaded processors will be moved.  This
continues until the load ratio between the most-to-least loaded
processor is below 1.05 or until no suitable grid can be found to
transfer.  A second load balancing option uses a space-filling Hilbert
curve to order the grids by their approximate spatial position.  Then,
once the grids have a specific one-dimensional ordering, we can divide
up the grids into $N_p$ groups (with the division taking place as
equally as possible).  Load balancing is done separately for each
level.  Clearly load balancing is most successful if there are
significantly more grids than processors; however, small grids are
less efficient because of the large number of ghost zones compared to
active zones, and so the code uses a
simple heuristic in order to split up grids until there are of order
10 grids per processor.  This generally results in good load-balancing
while not producing grids that are wastefully small.

Communication between processors is done using a non-blocking
communication strategy that allows overlap of communication and
computation.  This can be done efficiently because each processor
retains a copy of the entire hierarchy of grids, except that grids
that do not `live' on a given processor only contain meta data
(essentially just the location and size of the grid; such grids are denoted as
`ghost' grids).  Replicating the hierarchy means that all
communication of data from one grid to another can be identified by
each processor independently.  The metadata for `ghost' grids are
quite small and so the extra memory required is generally not onerous
unless very large numbers of grids are used (more than a few hundred
thousand grids).  A schematic of this distribution is shown in
Figure~\ref{fig.amr_hierarchy}.

Data is transferred through a three-step procedure that takes
advantage of the capabilities of the MPI library: (i) as the code
progresses and data is needed from another grid on another processor,
the receiving processor posts an MPI non-blocking receive indicating
that it is expecting data; this outstanding receive is recorded in a
table, (ii) the sending processor calls the MPI non-blocking send
function, and then finally (iii) the receiving processor, after it has
carried out all the computation it can, waits for any MPI message to
arrive.  Each message is coded so that it can be matched with the
appropriate receive posted in the first step, and based on that, the
appropriate routine is called to processes the data.  Step (iii) is
repeated until there are no outstanding receives.

@@

At each complete root grid timestep (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers of that name across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells updates per second per MPI process.  This
information is then 
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with further discussion of the
performance measurement implementation.

@@

It is advantageous to use \enzo\ in its ``unigrid'' (i.e.,
non-adaptive mesh) mode for a variety of problems, including
non-gravitating turbulence
\citep[e.g.,][]{2002ApJ...569L.127K,Kritsuk04, 2007ApJ...665..416K,
2009ASPC..406...15K}, the Lyman-$\alpha$ forest
\citep{2005MNRAS.361...70J,2009MNRAS.399.1934P}, or feedback of
metal-enriched gas from galaxies into the intergalactic medium
\citep{2004ApJ...601L.115N,2011ApJ...731....6S}.  Achieving good
scaling of the code in unigrid mode is relatively straightforward --
upon initialization, unigrid simulations are decomposed such that each
MPI process has a roughly equal subvolume (and thus an equal number of grid
cells), meaning that work is evenly distributed among computational elements.  Communication
patterns for both the gravity solve (which uses a fast Fourier
transform) and the fluid solves (which transfer boundary information
between subvolumes) are predictable and straightforward, and
rebuilding of the grid hierarchy does not take place, removing a
substantial global operation and a great deal of communication.

@@

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this value has
been empirically determined to be roughly $128^3$ cells per core.  If
fewer cells per core are used, the CPU is essentially data-starved 
and poor scaling is observed due to computing units being idle while
waiting for information to be communicated from other processes (for,
e.g., boundary information or gravity solves).  Substantially larger
cell counts per core would in principle help scaling by reducing the
amount of inter-process communication needed, but larger cell counts
are typically impractical on most machines due to the amount of memory
available on each node.

As a final point, we observe that scaling at larger core counts has
been measured, but only with an experimental hybrid-parallel (MPI +
OpenMP) version of \enzo.  Using this version, scaling comparable to
that shown in Figure~\ref{fig.uniscale} was seen on up to 98,304 cores
on the NICS Kraken XT5 (an upgraded version of the XT4 machine used
for the scaling study shown in the figure), using 2-8 OpenMP threads
per MPI process.  Hybrid parallelism has the potential to
substantially improve scaling by reducing the amount of communication
per grid tile, as described in the previous paragraph; however, the
optimal ratio of OpenMP threads per MPI task seems to vary
substantially between computational platforms and astrophysical
problems/required \enzo\ physics modules, and thus we hesitate to
provide strict guidelines here.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
\enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in \enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (the multi-grid
solver in AMR subgrids), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with fewer
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved more quickly, and the
opposite occurs with larger simulations.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per timestep is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late physical
time in the simulation.
Therefore, simply performing a few steps at the beginning of the
calculation does not produce a good estimate of the required CPU time.

@@

The physical reason for the $f_{l+1}/f_l$ ratios found in the
cosmological simulations appears to be due to a combination of the
density structure of individual clouds and to the distribution of the
clouds themselves.  In particular, we note that for a $\rho \propto
r^{-2}$ density profile, the Lagrangian refinement criteria typically used in
such simulations produces an $f_{l+1}/f_l$ ratio of 1/8 for a single,
resolved cloud.  Of course, at some point we would resolve the flat
density of individual clouds and the ratio would climb, making further
refinement more costly; however, we are not yet in this regime in this
example.

@@

Many of today's leading supercomputers use a heterogeneous
computing platform:
on a single node of a distributed-memory platform, a multi-core CPU is
often paired with one or more many-core accelerators.  One programming
model that has been shown to successfully take advantage of these
hardware accelerators is to run the serial component of the algorithm
on the CPU, and the vector-parallel part of the algorithm on the
many-core hardware accelerator. Due to the much higher computational
performance of the many-core accelerators, if a code can be ported to
effectively utilize this heterogeneous architecture, massive speedup
in simulation performance may be achieved.  This heterogeneous trend
is likely to continue in future supercomputers, due to the relatively
low energy needs of hardware accelerators as well as other
factors. Thus, to ensure \enzo's efficiency on future high-end
computational platforms, some of the most time-consuming parts of
\enzo\ have been ported to many-core architecture.  Since GPUs are
currently the most popular many-core accelerator, NVIDIA's CUDA C/C++
programming technology was chosen to port some \enzo\ modules to
NVIDIA GPUs.

@@

The key step in porting to many-core architectures such as the GPU is
exposing the massive parallelism inherent in the algorithm. Due to the
explicit, directionally-split stencil pattern of both the PPM and
Dedner MHD solvers, they are inherently massively parallel and thus
should be a good fit for hardware acceleration. Both solvers essentially
contain two parts -- computation of fluxes, and a cell update.
When computing fluxes, the basic procedure is to compute the flux at
each cell-interface given the inputs from neighboring cells. In the
cell update part, the fundamental computation is updating the
cell-centered values using the previously-computed fluxes. In the CPU
serial code, a loop over the grid is used, where the loop body
contains these basic computations. In both parts,
algorithmically-different loop iterations are completely independent
of each other. Thus, the natural parallelization scheme is to map
one GPU thread to one iteration of the loop. However, the original CPU
code, which operates on each grid as a serial process, does not
completely expose this parallelism as some small temporary arrays are
reused among loop iterations.  This re-use of arrays introduces data
dependency among loop iterations, which is undesirable for GPU
parallelization.  Because of this re-use of arrays, the main change in
porting the serial CPU solvers to allow massively parallel GPU
computation was replacing those shared temporary arrays by larger
temporary arrays that are not shared among loop iterations. This
change exposed the massive parallelism in the algorithm, which could
then be accelerated in a straightforward manner using CUDA.

To illustrate the speedup provided by porting two solvers to GPUs, we
show the results of a weak scaling test of driven MHD turbulence on a
uniform mesh in Figure~\ref{fig:gpu_scaling}.  This particular problem
type contains no physics other than the equations of ideal
MHD, and thus is representative of the type of calculation than can currently
benefit from the GPU-optimized solvers in \enzo.  In this scaling
test, we use the Dedner MHD solver, which runs on both CPUs and GPUs
and has been tuned to maximize performance on both platforms.  The
benchmarking platform is a Cray XK6 system with a Gemini
interconnect. Each node has a 16-core AMD Opteron 6272 CPU and a
single NVIDIA Tesla K20 GPU, which has 2,496 CUDA cores and a maximum
theoretical speed of 1.17 Tflops for double-precision computations.  For
the CPU run, all the 16 cores in each node are used by launching 16
MPI processes per node. For the GPU run, 16 MPI processes are also
launched per node, and all processes on this node share the single GPU
on that node. This can work because NVIDIA's MPS (Multi-Process
Service) technology allows multiple processes to concurrently use the
same GPU.  A series of weak scaling tests were run, varying from 1 to
8 nodes (16 to 128 MPI processes), with each node containing a cube of
$256^3$ cells (so, a 4-node computation would have a domain that is
$512 \times 512 \times 256$ cells).  Figure~\ref{fig:gpu_scaling}
displays the scaling results in terms of cell updates per second per
node -- as a result, ideal weak scaling is a horizontal line.  In
general, the simulations using the GPU-accelerated MHD solver perform
roughly 5 times better overall on this system than the equivalent
CPU-only calculation run on the same number of nodes.  We note,
however, that this somewhat mis-represents the speedup obtained by
porting the MHD code to GPUs -- in the GPU simulation case, the 16 CPU
cores on each node are mostly idle, as they are used only for
timestep calculation and boundary condition transmission (which are
both computationally inexpensive compared to the fluid solve).  This
means that, effectively, the GPU simulation is 80 times faster than a
single CPU core, so a user would likely see much greater effective
speedup on a system where the ratio of CPU cores to GPUs is lower.
